---
title: "Biostat 212a Homework 2"
subtitle: "Due Feb 11, 2026 @ 11:59PM"
author: "YOUR NAME and UID"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 4.8.1 (10pts)
Using a little bit of algebra, prove that (4.2) is equivalent to (4.3). In
other words, the logistic function representation and logit represen-
tation for the logistic regression model are equivalent.

logistic func: 
p(X) = e^(β0+β1X) / [1 + e^(β0+β1X)]

logit func:
p(X) / [1 - p(X)] = e^(β0 + β1X)
log(p(X) / [1 - p(X)]) = β0 + β1X

## ISL Exercise 4.8.6 (10pts)

Suppose we collect data for a group of students in a statistics class
with variables X1 = hours studied, X2 = undergrad GPA, and Y =
receive an A. We fit a logistic regression and produce estimated
coefficient, βˆ0 = −6, βˆ1 = 0.05, βˆ2 = 1.

(a) Estimate the probability that a student who studies for 40 h and
has an undergrad GPA of 3.5 gets an A in the class.
(b) How many hours would the student in part (a) need to study to
have a 50 % chance of getting an A in the class?

## ISL Exercise 4.8.9 (10pts)

This problem has to do with odds.
(a) On average, what fraction of people with an odds of 0.37 of
defaulting on their credit card payment will in fact default?
(b) Suppose that an individual has a 16 % chance of defaulting on 
her credit card payment. What are the odds that she will de-
fault?

## ISL Exercise 4.8.13 (a)-(i) (50pts) [applied question]

13. This question should be answered using the Weekly data set, which
is part of the ISLR2 package. This data is similar in nature to the
Smarket data from this chapter’s lab, except that it contains 1, 089
weekly returns for 21 years, from the beginning of 1990 to the end of
2010.

```{r}
library(ISLR2)
library(tidyverse)
library(ggplot2)
```

```{r}
# load the data and take a look at the first few rows
ISLR2::Weekly %>% head()
```

(a) Produce some numerical and graphical summaries of the Weekly
data. Do there appear to be any patterns?

```{r}
# numerical summaries
Weekly %>%  summary()
```
```{r}
# graphical summaries; leave out the Direction variable for now
pair(Weekly[, -9], pch = 20, cex.labels = 0.7, font.labels = 1) # pairwise scatterplots of the predictors

# manually
Weekly %>%  ggplot(aes(x = Lag1, y = Lag2, color = Direction)) +
  geom_point() +
  labs(title = "Lag1 vs Lag2 colored by Direction")
```
```{r}
library(GGally)
ggpairs(Weekly[, -9], aes(color = Direction)) # pairwise scatterplots with color by Direction
```

(b) Use the full data set to perform a logistic regression with
Direction as the response and the five lag variables plus Volume
as predictors. Use the summary function to print the results. Do
any of the predictors appear to be statistically significant? If so,
which ones?

```{r}
# fitting logistic regression model with Direction as response and Lag1, Lag2, Lag3, Lag4, Lag5, Volume as predictors
logistic_model <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = Weekly, family = binomial)
summary(logistic_model)
```

The predictors that appear to be statistically significant are___


(c) Compute the confusion matrix and overall fraction of correct
predictions. Explain what the confusion matrix is telling you
about the types of mistakes made by logistic regression.

https://www.digitalocean.com/community/tutorials/confusion-matrix-in-r
```{r}
library(caret)
# predict the probabilities of the positive class (Up) using the logistic regression model
confusion_matrix <- confusionMatrix(data = predict(logistic_model, type = "response") > 0.5, reference = Weekly$Direction)

# accyuracy of the model (overall fraction of correct predictions)
model_accuracy <- confusion_matrix$overall['Accuracy']
model_accuracy_manual <- sum(diag(confusion_matrix$table)) / sum(confusion_matrix$table)
```

(d) Now fit the logistic regression model using a training data period
from 1990 to 2008, with Lag2 as the only predictor. Compute the
confusion matrix and the overall fraction of correct predictions
for the held out data (that is, the data from 2009 and 2010).

* Lag2 as the only predictor
* training data period from 1990 to 2008
* held out data from 2009 and 2010


(e) Repeat (d) using LDA.
(f) Repeat (d) using QDA.
(g) Repeat (d) using KNN with K = 1.
(h) Repeat (d) using naive Bayes.

(i) Which of these methods appears to provide the best results on
this data?

The method that appears to provide the best results on this data is___

## Bonus question: ISL Exercise 4.8.13 Part (j) (30pts)

(j) Experiment with different combinations of predictors, includ-
ing possible transformations and interactions, for each of the

methods. Report the variables, method, and associated confu-
sion matrix that appears to provide the best results on the held

out data. Note that you should also experiment with values for
K in the KNN classifier.

## Bonus question: ISL Exercise 4.8.4 (30pts)

When the number of features p is large, there tends to be a deteri-
oration in the performance of KNN and other local approaches that
perform prediction using only observations that are near the test ob-
servation for which a prediction must be made. 

This phenomenon isknown as the curse of dimensionality, and it ties into the fact that non-parametric approaches 
often perform poorly when p is large. We will now investigate this curse.

(a) Suppose that we have a set of observations, each with measure-
ments on p = 1 feature, X. We assume that X is uniformly

(evenly) distributed on [0, 1]. Associated with each observation

is a response value. Suppose that we wish to predict a test obser-
vation’s response using only observations that are within 10 % of

the range of X closest to that test observation. For instance, in
order to predict the response for a test observation with X = 0.6,
we will use observations in the range [0.55, 0.65]. On average,
what fraction of the available observations will we use to make
the prediction?
(b) Now suppose that we have a set of observations, each with
measurements on p = 2 features, X1 and X2. We assume that
(X1, X2) are uniformly distributed on [0, 1] × [0, 1]. We wish to
predict a test observation’s response using only observations that
are within 10 % of the range of X1 and within 10 % of the range
of X2 closest to that test observation. For instance, in order to
predict the response for a test observation with X1 = 0.6 and
X2 = 0.35, we will use observations in the range [0.55, 0.65] for
X1 and in the range [0.3, 0.4] for X2. On average, what fraction
of the available observations will we use to make the prediction?

(c) Now suppose that we have a set of observations on p = 100 fea-
tures. Again the observations are uniformly distributed on each

feature, and again each feature ranges in value from 0 to 1. We
wish to predict a test observation’s response using observations
within the 10 % of each feature’s range that is closest to that test
observation. What fraction of the available observations will we
use to make the prediction?
(d) Using your answers to parts (a)–(c), argue that a drawback of

KNN when p is large is that there are very few training obser-
vations “near” any given test observation.

(e) Now suppose that we wish to make a prediction for a test obser-
vation by creating a p-dimensional hypercube centered around

the test observation that contains, on average, 10 % of the train-
ing observations. For p = 1, 2, and 100, what is the length of

each side of the hypercube? Comment on your answer.

Note: A hypercube is a generalization of a cube to an arbitrary
number of dimensions. When p = 1, a hypercube is simply a line
segment, when p = 2 it is a square, and when p = 100 it is a
100-dimensional cube.